An Analysis of the Feasibility and Performance of BitTorrent as a Data Transfer Protocol in Peer to Peer Backup Systems

### 2014-04-24 paper/paper.tex
Paper: redo results for filesize vs numfiles scenario

In this scenario, we tested how our system handles a constant amount of data spread out across a varying number of files. The goal of this test is twofold. First, it allows us to test the capabilities of BitTorrent Sync in handling a large number of backups. From the data gathered here, it is possible to determine how well BitTorrent Sync would scale for use in a production system, where thousands of users could each be backing up thousands of files of varying sizes. Second, it allows us to measure the performance of BitTorrent against different file sizes. Since the use case for BitTorrent is typically downloading large files, it is important to determine whether or not our system will perform well when users try to backup many small files, should it be used in a production environment.

% TODO: Maybe we should not mention the 1MB test at all.
Our tests for this scenario used 8 virtual machines in SZTAKI Cloud: 1 peer, 5 nodes, 1 tracker, and 1 syslog-ng server. To have a wide range of file sizes, we tested with 10MB, 100MB, and 1GB files. In each test instance, the total amount of data backed up was 1GB, so 100 files, 10 files, and 1 file were backed up, respectively. The procedure for each test instance is as follows:

@@

Figure \ref{fig:PeerRecoverySpeed} shows the speed of a peer recovering 1 GB of data in three different situations. Each situation differs in how many files comprise the 1 GB payload: one 1 GB file, ten 100 MB files, and one hundred 10 MB files. The moving average of each data series is shown instead of the raw data to reduce visual clutter; the trends are analogous.

\begin{figure}
  \centerline{\includegraphics[scale=.8]{figures/PeerRecoverySpeed}}
  \caption{The speed at which a peer recovered 1 GB worth of data. \label{fig:PeerRecoverySpeed}}
\end{figure}

@@

\begin{table}
\begin{center}
    \begin{tabular}{| l | l | l | l |}
    \hline
    & One 1 GB File & Ten 100 MB files & Hundred 10 MB files \\ \hline
    Mean & 8318.52 & 7210.29 & 5405.28 \\ \hline
    Standard Deviation & 3009.26 & 3444.88 & 1738.38 \\ \hline
    Median & 9470.31 & 8599.82 & 5923.23\\ \hline
    Max & 11691.1 & 10926.5 & 8290.26\\ \hline
    \end{tabular}
    \caption{Statistics on Peer Data Recovery Speed (in kB/s)}
    \label{tab:PeerRecoverySpeed}
\end{center}
\end{table}

It is worth noting that the increase in speed does not scale linearly with the size of the file. Although the size of the files increased by the same order of magnitude between the trials, the increase in recovery speed seems to have diminishing returns. Table \ref{tab:PeerRecoverySpeed} provides relevant statistics. The mean recovery speed increased by 33\% between the "Hundred 10 MB Files" trial and the "Ten 100 MB Files" trial. Compare this to an increase of 15\% between the "Ten 100 MB Files" trial and the "One 1 GB File" trial. Furthermore, the total time it took to recover the file between the former trials improves by 60 seconds, whereas the recovery time between the latter trials improves by only 3 seconds.

@@

Figure \ref{fig:NodeRecoverySpeed} shows the combined speed at which the nodes are uploading a 1 GB file to a peer. A moving average is used to reduce visual clutter. We can see a significant improvement between the combined upload speed achieved by all nodes compared to the individual upload speeds of each node.

\begin{figure}
  \centerline{\includegraphics[scale=.8]{figures/NodeRecoverySpeed}}
  \caption{The combined speed of nodes uploading 1 GB worth of data to a peer. \label{fig:NodeRecoverySpeed}}
\end{figure}

If the peer was restricted to recovering its data from only one node, its download speed would be limited by the node's upload speed. Table \ref{tab:NodeRecoverySpeed} lists the relevant upload speed statistics of the nodes. At best, the peer could download at an average of 2813 kB/s. However, by using BitTorrent as a data transfer protocol instead of a direct download, the peer's download speed is limited by the \textit{sum} of each node's upload speed: 7981 kB/s. This means, on average, the peer can theoretically download 280\% faster!\footnote{This number is based on the mean. The median suggests that the peer can theoretically download 207\% faster.}

\begin{table}
\begin{center}
    \begin{tabular}{| l | l | l | l | l | l |}
    \hline
    & Node 1 & Node 2 & Node 3 & Node 4 & Node 5 \\ \hline
    Mean & 2813.57 & 1441.16 & 1780.92 & 1417.63 & 530.91 \\ \hline
    Standard Deviation & 2316.67 & 1765.76 & 1758.64 & 1837.77 & 913.80 \\ \hline
    Median & 2281.84 & 703.98 & 1169.43 & 384.91 & 195.90 \\ \hline
    Max & 9000.08 & 7516.42 & 7600.65 & 7071.78 & 5147.33 \\ \hline
    \end{tabular}
    \caption{Statistics on Node Upload Speed (in kB/s)}
    \label{tab:NodeRecoverySpeed}
\end{center}
\end{table}

@@

Figure \ref{fig:EntityBackupSpeed} shows the speeds at which each participating entity is backing up a 1 GB file.  We use a 20 period moving average to reduce the visual clutter while keeping the overall trends the same.\footnote{The use of a 20 period moving average significantly alters the peaks and valleys of each line. This is acceptable for our purposes because we are only concerned with their existence and relative placements, not their individual details.}

\begin{figure}
  \centerline{\includegraphics[scale=.8]{figures/EntityBackupSpeed}}
  \caption{The speed at which each entity backs up a 1 GB file. \label{fig:EntityBackupSpeed}}
\end{figure}

We confirm that the nodes, in addition to the peer, are actively backing up the peer's file. This is the expected behavior of the BitTorrent protocol, which allows all users in a swarm to share data with each other. BitTorrent Sync, which uses a modified BitTorrent protocol, also seems to leverage this feature. % TODO why is the peer consistently uploading faster than nodes? seeder vs leecher maybe?

We believe the initial ~30 seconds of inactivity is due to BitTorrent Sync indexing the 1 GB file before it can transfer it.

@@

>>>>

In this scenario, we tested how our system handles a constant amount of data spread out across a varying number of files. We run this test for three reasons. First, it allows us to test the capabilities of BitTorrent Sync in handling a large number of backups. From the data gathered here, it is possible to determine how well BitTorrent Sync would scale for use in a production system, where thousands of users could each be backing up thousands of files of varying sizes. Second, it allows us to measure the performance of BitTorrent against different file sizes. Since the use case for BitTorrent is typically downloading large files, it is important to determine whether or not our system will perform well when users try to backup many small files, should it be used in a production environment. Third, it will demonstrate the ability of our system to backup and recover files faster than if they were transferred by a direct download protocol, as opposed to BitTorrent.

Our tests for this scenario used 8 virtual machines in SZTAKI Cloud: 1 peer, 5 nodes, 1 tracker, and 1 syslog-ng server. To have a wide range of file sizes, we tested with 10MB, 100MB, and 1GB files. In each test instance, the total amount of data backed up was 1GB, so 100 files, 10 files, and 1 file were backed up, respectively. The procedure for each test instance is as follows:

@@

Figure \ref{fig:PeerRecoverySpeed} shows the speed of a peer recovering 1 GB of data in three different situations. Each situation differs in how many files comprise the 1 GB payload: one 1 GB file, ten 100 MB files, and one hundred 10 MB files. The moving average of each data series is shown instead of the raw data to reduce visual clutter; the trends are analogous. The usage of moving averages is the reason why some lines do not cross the x-axis.

\begin{figure}
  \centerline{\includegraphics[scale=1]{figures/PeerRecoverySpeed}}
  \caption{The speed at which a peer recovered 1 GB worth of data. \label{fig:PeerRecoverySpeed}}
\end{figure}

@@

\begin{table}
\begin{center}
    \begin{tabular}{| l | l | l | l |}
    \hline
    & One 1 GB File & Ten 100 MB files & Hundred 10 MB files \\ \hline
    Mean & 8144.84 & 7874.69 & 5559.32 \\ \hline
    Standard Deviation & 2564.97 & 1899.86 & 1649.70 \\ \hline
    Median & 9121.41 & 8379.84 & 5874.48 \\ \hline
    Max & 11478.79 & 10948.98 & 8242.56 \\ \hline
    Duration (sec) & 162 & 165 & 231 \\ \hline
    \end{tabular}
    \caption{Statistics on Peer Data Recovery Speed (in kB/s)}
    \label{tab:PeerRecoverySpeed}
\end{center}
\end{table}

It is worth noting that the increase in speed does not scale linearly with the size of the file. Although the size of the files increased by the same order of magnitude between the trials, the increase in recovery speed seems to have diminishing returns. Table \ref{tab:PeerRecoverySpeed} provides relevant statistics. The mean recovery speed increased by 41.65\% between the "Hundred 10 MB Files" trial and the "Ten 100 MB Files" trial. Compare this to an increase of 3.43\% between the "Ten 100 MB Files" trial and the "One 1 GB File" trial. Looking at it another way, the total time it took to recover the file between the former trials improves by 40\%, whereas the recovery time between the latter trials improves by only 4.76\%.

@@

Figure \ref{fig:NodeRecoverySpeed} shows the combined speed at which the nodes are uploading a 1 GB file to a peer during the file recovery process. A moving average is used to reduce visual clutter. We can see a significant improvement between the combined upload speed achieved by all nodes compared to the individual upload speeds of each node.

\begin{figure}
  \centerline{\includegraphics[scale=1]{figures/NodeRecoverySpeed}}
  \caption{The combined speed of nodes uploading a 1 GB file to a peer for file recovery. \label{fig:NodeRecoverySpeed}}
\end{figure}

% TODO analyze graph? 10sec inactivity in beginning, node 2 doesn't start uploading until 35secs, node 5 barely uploads, duration, other characteristics

If the peer was restricted to recovering its data from only one node, its download speed would be limited by the node's upload speed. Table \ref{tab:NodeRecoverySpeed} lists the relevant upload speed statistics of the nodes. At best, the peer could download at an average of 2303.07 kB/s. However, by using BitTorrent as a data transfer protocol instead of a direct download, the peer's download speed is limited by the \textit{sum} of each node's upload speed: 7427.61 kB/s. This means, on average, the peer can theoretically download 322.50\% faster!\footnote{This number is based on the mean. The highest median is 2024.57 kB/s and the sum of medians is 4889.01 kB/s, 241.48\% faster.}

\begin{table}
\begin{center}
    \begin{tabular}{| l | l | l | l | l | l |}
    \hline
    & Node 1 & Node 2 & Node 3 & Node 4 & Node 5 \\ \hline
    Mean & 2303.07 & 1293.71 & 1702.47 & 1556.95 & 571.41 \\ \hline
    Standard Deviation & 1921.11 & 1714.13 & 1852.66 & 1796.30 & 1000.72 \\ \hline
    Median & 2024.57 & 477.37 & 1187.60 & 1028.09 & 171.38 \\ \hline
    Max & 7632.07 & 7423.9 & 6936.94 & 7237.46 & 5216.15 \\ \hline
    \end{tabular}
    \caption{Statistics on Node Upload Speed (in kB/s)}
    \label{tab:NodeRecoverySpeed}
\end{center}
\end{table}

@@

Figure \ref{fig:EntityBackupSpeed} shows the speeds at which each participating entity is backing up a 1 GB file.  We use a 20 period moving average to reduce the visual clutter while keeping the overall trends the same.\footnote{The use of a 20 period moving average significantly alters the peaks and valleys of each line. This is acceptable for our purposes because we are only concerned with each line's existence and relative placement, not its individual details.}

\begin{figure}
  \centerline{\includegraphics[scale=1]{figures/EntityBackupSpeed}}
  \caption{The speed at which each entity backs up a 1 GB file. \label{fig:EntityBackupSpeed}}
\end{figure}

We confirm that the nodes, in addition to the peer, are actively backing up the peer's file. This is the expected behavior of the BitTorrent protocol, which allows all users in a swarm to share data with each other. BitTorrent Sync, which uses a modified BitTorrent protocol, also seems to leverage this feature.

The peer generally uploads the fastest for the first 9 minutes; the nodes are all requesting chunks from mostly the peer throughout this period, as it is the initial seeder. However, around the 9 minute mark, the nodes seem to start requesting data from each other more often than the peer, as seen by the peer's decrease in upload speed between the 9 and 10 minute mark.

% TODO what is happening after the 11 minute mark? why are they all uploading again? it might help to look at the download speed graph of everyone

We believe the initial ~40 seconds of inactivity is due to BitTorrent Sync indexing the 1 GB file before it can transfer it.

\subsection{Conclusion}

We have seen BitTorrent Sync handle 
% 1. how well does btsync handle large number backups? how well will it scale?
% 2. measure performance of bittorrent against diff file sizes
% 3. bittorrent faster than direct download?

@@

% summarize contributions and impact
